Outline flowCore paper

### 2008-05-09 paper/manuscript.tex
git-svn-id: https://hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@31536 bc3139a8-67e5-0310-9ffc-ced21a209358
Perry's updates to the introduction and conclusions.
git-svn-id: https://hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@31600 bc3139a8-67e5-0310-9ffc-ced21a209358

\section*{Introduction}
Traditionally, flow cytometry (FCM) has been a tube-based technique
limited to small-scale laboratory and clinical studies.  High
throughput methods for flow cytometry high content screening (FC-HCS)
have recently been developed and are now used in both basic and
clinical research. As a result, the need for flexible and well
structured computational tools to efficiently handle and analyze
FC-HCS has increased considerably.
However, the absence of a shared research platform that
bioinformaticians, computer scientists, and statisticians can collaboratively
use has hindered the development of novel methods 
for FCM data analysis \citep{lizard2007fca}.
We believe that the open source statistical
software R, in conjunction with the Bioconductor Project can fill this
gap.  
%% EAS: I think we mention flowCore before talking about data structures and
% methods.  We should first say explicity that we have a solution for
% the missing shared platform.
We have developed the Bioconductor package flowCore as a freely available,
sophisticated and extensible FCM data analysis platform that enables
researchers to efficiently handle FC-HCS data and encourages open
development of tools for their coherent analysis. 
%In this paper, we propose R data structures to handle FCM data
%through the main steps of importing, storing, assessing and
%preprocessing. 
%These structures, along with a range of specialized
%methods for compensation, transformation, gating and additional
%preprocessing steps, are implemented in the newly developed
%Bioconductor package flowCore.  flowCore is a freely available,
%sophisticated and extensible FCM data analysis platform that enables
%researchers to efficiently handle FC-HCS data and enables open
%development of new methods and tools for their coherent
%analysis.
flowCore runs on Windows, Mac OS X and Linux/Unix operating
systems.

\subsection*{R and Bioconductor}
R is a robust statistical programing environment which offers a wide
range of statistical and visualization methods developed for various
fields of application. In addition, R is an open source research %already mentioned R, and by extention flowCore open source
platform that bioinformaticians, computer scientists, and
statisticians routinely use to develop novel tools and methodology for
statistical data analysis. For biological applications, the
Bioconductor project has become the standard toolset
\citep{gentleman2006bos} to process data spanning a diversity of
research fields from genomics to proteomics to complex cell
biology. It is particularly geared towards the analysis of large
numeric datasets that typically arise from high throughput experiments
such as microarray gene expression analysis or FC-HCS. A shared
platform proved to be particularly beneficial for the development of
analysis routines for microarray gene expression data, and by today
many of these early innovations have been incorporated into commercial
software products, thus greatly improving the quality of gene
expression data analysis. 
%% NLM: that's probably the important point: platform for innovation rather than for standardization
Shared data structures have been shown to be
essential for such a collaborative effort. We envision a similar
paradigm for FC-HCS data.

@@

The flowCore framework presented here offers import functionality for
raw data FCS files along with their complete set of file-specific meta
information (Figure~\ref{fig1:FrameWork}). Moreover, it is a software
implementation of the Gating-ML CR, which opens the possibility to
integrate flowCore in existing workflows and to communicate with
every other FCM tool that adheres to the proposed
standard.

flowCore is not a GUI-driven software, and all operations are done
using a command line interface. Conceptually, it is possible to add a
more elaborate user interface on top of that, however the focus is on
a programmatic approach to enable the convenient development of novel
analysis methods. By taking the burden of data management from the
programmer, and by providing well-defined APIs and a structured class
hierarchy, it is possible to readily test new ideas and to easily
extend the frameworks functionality. %Both the underlying R and
%Bioconductor software and flowCore itself are open source and open %already mentioned its open source
%development projects, thus community involvement in the developing
%process is highly appreciated.

@@

\subsection*{Flow Set}
In high-throughput FCM, many of the analysis tasks need to be
performed consistently across multiple flow measurements, hence we
introduce the concept of a collection of flow frames called a flow
set. Flow sets store the relevant information associated with each
individual frame such as descriptions of the samples or the treatment
to which a sample was subjected, and they manage the consistent
application of operations on the individual flow frames. The burden of
keeping score of the annotation is shifted from the user to the
infrastructure, thus reducing the risk of common errors % (\textit{i.e.}, example of an error ) 
that occur when using more or less ad hoc spreadsheets that
lack a controlled structure. There is no native ordering
to a flow set, which in a way reflects the still predominant
application of tube-based experiments rather than experiments
following a plate-design. 
%% EAS: I don't think the previous sentence is required. Does it 
%% matter if the flowSet is ordered or not?

@@

Through flowCore, we have provided the FCM community a freely
available, sophisticated, development and analysis platform for high
throughput data analysis. The scope and complexity of the technology
is evolving rapidly, and a collaborative effort in devising new
methodology has been proven beneficial for a number of different
biological and computational biology challenges. We hope that our
framework will be the foundation for fruitful shared research by many
collaborators from multiple scientific fields, further pushing the
boundaries of the exiting and still promising FCM technology.

>>>>

\section*{Introduction}
During the last several years, automation technologies have been developed
that enable the use of flow cytometry high content screening (FC-HCS)
to generate large, complex datasets in both basic and clinical research
applications \citep{references from Ryan's FC-HCS paper}. A
serious bottleneck in the intepretation of existing studies and the application of
FC-HCS to even larger, more complex problems is that data management
and data analysis methods have not advanced sufficiently far from the
methods developed for 
traditional applications of flow cytometry (FCM) to small-scale, tube-based
studies.

Some of the consequences of this lag in development are difficulties in 
maintaining the integrity and documentation of extremely large datasets,
assessing measurement quality, developing validated assays, assessing the
accuracy of gating techniques, automating complex gating strategies, and
aggregating statistical results across large study sets for further analysis.
Among the most serious consequences of the current situation is, however, 
that it is very difficult for new analysis
approaches to find there way into standard practice. We believe that
this barrier to the development and dissemination of new analysis methods
is one of the fundamental restraints on the
future expansion of FC-HCS in both clinical and research applications.

In this paper, we describe a set of flexible and well
structured computational tools to efficiently analyze
FC-HCS. This set of tools is based on the open source statistical
software R, in conjunction with the Bioconductor Project. Our intent
is to provide a shared research platform that enables
bioinformaticians, computer scientists, and statisticians to work
collaboratively with biologists and clinicians to develop novel methods \citep{lizard2007fca} for FCM
data analysis. A critical component of this approach is the development and implementation
of standards that will facilitate the adoption of these methods
by both the larger research community and commercial software vendors.
Consequently, we are optimistic that this software platform will play a
critical role in addressing the bottlenecks described above that limit
further advances in the application of FC-HCS.

%% EAS: I think we mention flowCore before talking about data structures and
% methods.  We should first say explicity that we have a solution for
% the missing shared platform.
The computational tools we have developed are distributed in the R 
software language (www.r-project.org) as the Bioconductor
(www.bionductor.org) package flowCore. flowCore is a freely available, highly
functional and extensible FCM data analysis platform that 
enables researchers to efficiently handle FC-HCS data and encourages open
development of tools for their coherent analysis. Among the issues
addressed by flowCore are computationally efficient data structures and
a range of specialized methods for compensation, transformation, and gating. 
flowCore runs on Windows, Mac OS X and Linux/Unix operating
systems.  In this paper, we describe the functionality and methods underlying flowCore, 
and provide examples of their use.

\subsection*{R and Bioconductor}
R is a robust statistical programing environment which offers a wide
range of statistical and visualization methods developed for various
fields of application. In addition, R is an open source research %already mentioned R, and by extention flowCore open source
platform that bioinformaticians, computer scientists, and
statisticians routinely use to develop novel tools and methodology for
statistical data analysis. For biological applications, the
Bioconductor project has become the standard toolset
\citep{gentleman2006bos} to process data spanning a diversity of
research fields from genomics to proteomics to cell
biology. It is particularly geared towards the analysis of large
numeric datasets that typically arise from high throughput experiments
such as microarray gene expression analysis or FC-HCS. A shared
platform has proven to be particularly beneficial for the development of
analysis routines for microarray gene expression data, and by today
many of these early innovations have been incorporated into commercial
software products, thus greatly improving the quality of gene
expression data analysis. 
%% NLM: that's probably the important point: platform for innovation rather than for standardization
Shared data structures have been shown to be
essential for such a collaborative effort. We envision a similar
paradigm for FC-HCS data.

@@

The flowCore framework presented here offers import functionality for
raw data FCS files along with their complete set of file-specific meta
information (Figure~\ref{fig1:FrameWork}). Moreover, it is a software
implementation of the Gating-ML CR, which opens the possibility to
integrate flowCore in existing workflows and to communicate with
every other FCM tool that adheres to the proposed
standard. Adherence to standards also plays a critical role in the ability
of new methods based on flowCore to find there way back into the
standard practices for FCM data analysis.

flowCore is not a GUI-driven software, and all operations are done
using a command line interface. Conceptually, it is possible to add a
more elaborate user interface on top of that, however the focus is on
a programmatic approach to enable the convenient development of novel
analysis methods and automation of complex analysis approaches. By taking the
burden of data management from the programmer, and by providing well-defined 
APIs and a structured class
hierarchy, it is possible to readily test new ideas and to easily
extend the framework's functionality. %Both the underlying R and
%Bioconductor software and flowCore itself are open source and open %already mentioned its open source
%development projects, thus community involvement in the developing
%process is highly appreciated.

@@

\subsection*{Flow Set}
In high-throughput FCM, many of the analysis tasks need to be
performed consistently across multiple flow measurements, hence we
introduce the concept of a collection of flow frames called a flow
set. Flow sets store the relevant information associated with each
individual frame such as descriptions of the samples or the treatment
to which a sample was subjected, and they manage the consistent
application of operations on the individual flow frames. The burden of
keeping score of the annotation is shifted from the user to the
infrastructure, thus reducing the risk of common errors % (\textit{i.e.}, example of an error ) 
that occur when using more or less ad hoc spreadsheets that
lack a controlled structure. There is no native ordering
to a flow set, which in a way reflects the still predominant
application of tube-based experiments rather than experiments
that depend on automated sample acquisition using a plate-based platform. 
%% EAS: I don't think the previous sentence is required. Does it 
%% matter if the flowSet is ordered or not?

@@

\secttion*{Conclusions}

Through flowCore, we have provided the FCM community a freely
available, highly functional, development and analysis platform for high
throughput data analysis. Our intent is to provide a platform
for the collaborative development of new analysis methods that
will facilitate the transition of these new methods to the larger 
flow community. Our experience has been that the 
collaborative effort we enable for devising new methodology has been proven
beneficial for a number of different biological and computational biology challenges. 
We hope that our
framework will be the foundation for fruitful shared research by many
collaborators from multiple scientific fields and will help resolve
bottlenecks that currently prevent the further development of and deployment of
FC-HCS to increasingly complex and important scientific and clinical 
applications.
